A weakly structured stem for human origins in Africa

### 2023-02-27 paper.tex
added refs to abstract

    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
    considerable uncertainty surrounds specific models of divergence and
    migration across the continent. Progress is hampered by limited fossil and
    genomic data, as well as variability in prior divergence time estimates.
    We use linkage disequilibrium and diversity-based statistics, optimized for
    rapid, complex demographic inference to discriminate among such models.  We
    infer detailed demographic models for populations across Africa, including
    representatives from eastern and western groups and newly sequenced whole
    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
    population history in which contemporary population structure dates back to
    Marine Isotope Stage (MIS) 5. The earliest population divergence among
    contemporary populations occurs 120--135ka and was preceded by structure
    between two or more weakly differentiated ancestral \emph{Homo} populations
    connected by gene flow over hundreds of thousands of years. Such \emph{weakly
    structured stem} models better explain patterns of
    polymorphism that were previously attributed to contributions from archaic
    hominins in Africa. In contrast to models with archaic introgression, we
    predict fossil remains from coexisting ancestral populations should be
    genetically and morphologically similar, and that an inferred 1--4\% of genetic
    differentiation among contemporary human populations can be attributed to
    genetic drift between stem populations. We show that model misspecification
    explains variation in previous divergence time estimates and argue that
    studying a suite of models is key to robust inferences about deep history.

>>>>

    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
    considerable uncertainty surrounds specific models of divergence and
    migration across the continent\citep{Henn2018-rf}. Progress is hampered by limited fossil and
    genomic data, as well as variability in prior divergence time estimates\citep{Henn2018-rf}.
    To discriminate among such models, we consider linkage disequilibrium and diversity-based 
    statistics, optimized for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We
    infer detailed demographic models for populations across Africa, including
    representatives from eastern and western groups and newly sequenced whole
    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
    population history in which contemporary population structure dates back to
    Marine Isotope Stage (MIS) 5. The earliest population divergence among
    contemporary populations occurs 120--135ka and was preceded by structure
    between two or more weakly differentiated ancestral \emph{Homo} populations
    connected by gene flow over hundreds of thousands of years. Such \emph{weakly
    structured stem} models better explain patterns of
    polymorphism that were previously attributed to contributions from archaic
    hominins in Africa \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic introgression, we
    predict fossil remains from coexisting ancestral populations should be
    genetically and morphologically similar, and that an inferred 1--4\% of genetic
    differentiation among contemporary human populations can be attributed to
    genetic drift between stem populations. We show that model misspecification
    explains variation in previous divergence time estimates and argue that
    studying a suite of models is key to robust inferences about deep history.
